The very dependent recursive structure of iterated parametricity in indexed form

### 2025-12-26 paper/paper.tex
Intro: Minor corrections + synthetic table

From the point of view of realisability, unary parametricity over a language is a realisability using the language itself as language of realisers. Like parametricity, realisability can be expressed in a syntactic way, that is as an interpretation from one language to another (i.e. mapping an expression $t$ of $\vdash A$ in some formal system to a derivation $t^*$ of $\vdash' |t| \real A$ in some other formal system, for some program $|t|$ extracted from $t$ -- namely $t$ itself in the case of parametricity), or a semantic way\footnote{This distinction between syntactic realisability and semantic realisability is itself connected to the difference between relative and absolute consistency. In the relative case, consistency of a formal system is reduced to the consistency of another formal system, and this can be done in a very weak logical system such as Primitive Recursive Arithmetic (PRA). On the other side, in the absolute case, consistency of a formal system is reduced to the consistency of the language of discourse, which, by G\"odel incompleteness theorem, if actually consistent, necessarily has to be as strong as the modelled formal system.}, that is as an interpretation in the language of discourse, which we could call in term the ``metalanguage'' (i.e. interpreting $A$ as a set $\textbf{A}$ or a relation $=_A$ of the language of discourse and interpreting a derivation $t$ of $\vdash A$ as an expression $|t|$ such that, respectively, $|t| \in \textbf{A}$ or $|t| =_A |t|$). Thus, the comparison between parametricity and realisability includes the case of reducibility~\cite{Tait}, which is a semantic form of realisability. In the same vein, binary parametricity over a language can be seen as defining logical relations~\cite{Plotkin} over itself, since logical relations are a binary form of reducibility.

@@

Parametricity, like realisability, reductibility, and logical relations, was most commonly handled in an indexed form, that is, using a type-theoretic language, considering predicates of type $A \rightarrow \Prop$, or families of type $A \rightarrow \MySet$, or relations of type $A \times A \rightarrow \Prop$, or relevant relations -- sometimes called \emph{correspondences} -- of type $A \times A \rightarrow \MySet$, where $\Prop$ and $\MySet$ are types representing, that is classifying -- in category theory terminology --, propositions or sets, respectively. Along the indexed-fibered correspondence, that is along the equivalence between the types $A \rightarrow \Prop$ and $\Sigma B:\MySet. (B \stackrel{inj}{\rightarrow} A)$ (proof-irrelevant case) and $A \rightarrow \MySet$ and $\Sigma B:\MySet. (B \rightarrow A)$ (proof-relevant case), for $A:\MySet$, everything told in the indexed form can morally also be told in fibered form. Such fibered approach to parametricity was for instance considered in Ghani \cite{Ghani...}, who also suggested to iterate parametricity.

@@

A type-theoretic analysis of parametricity in unary indexed form was obtained in \cite{Moulin}, together with an interpretation in a ``unary'' form of cubical set, thus implicitly in fibered form, in \cite{BernardyCoquandMoulin}. The connection between iterated binary parametricity in fibered form and cubical sets can also be found in Johann and Sojakova~\cite{JohannSojakova}. In particular, it became clear at this point that iterated binary parametricity produces semi-cubical sets while it was also observed\footnote{The authors learned it from Hugo Moeneclaye and Thorsten Altenkirch.} that iterated unary parametricity produces augmented semi-simplicial sets.

...

In a previous paper~\citep{hr25}, we exploited this correspondence to construct at the same time augmented semi-simplicial and semi-cubical sets in \emph{indexed form} as a special case of iterated parametricity. That is, instead of the following illustration of the cubical case (upto faces identities),

\begin{equation*}
  \begin{tikzcd}
    X_0: \U & X_1: \U \arrow[l, "\partial^L" description, shift left=2] \arrow[l, "\partial^R" description, shift right=2] & X_2: \U \arrow[l, "\partial^{L\kstar}" description, shift left=6] \arrow[l, "\partial^{R\kstar}" description, shift left=2] \arrow[l, "\partial^{\kstar L}" description, shift right=2] \arrow[l, "\partial^{\kstar R}" description, shift right=6] & \ldots
  \end{tikzcd}
\end{equation*}
we consider a family of families as shown below:
\begin{equation*}
  \begin{array}{lll}
    E_0 & :               \U                                                                          \\
    E_1 & :               E_0 \times X_0 \rightarrow  \U                                              \\
    E_2 & : \Pi a b c d.\,  E_1(a,b) \times E_1 (c,d) \times E_1(a,c) \times E_1 (b,d) \rightarrow \U \\
    \ldots
  \end{array}
\end{equation*}

>>>>

From the point of view of realisability, unary parametricity over a language is a realisability using the language itself as language of realisers. Like parametricity, realisability can be expressed in a \emph{syntactic} way, that is as an interpretation from one language to another (i.e. mapping an expression $t$ of $\vdash A$ in some formal system to a derivation $t^*$ of $\vdash' |t| \real A$ in some other formal system, for some program $|t|$ extracted from $t$ -- namely $t$ itself in the case of parametricity), or a \emph{semantic} way\footnote{This distinction between syntactic realisability and semantic realisability is itself connected to the difference between relative and absolute consistency. In the relative case, consistency of a formal system is reduced to the consistency of another formal system, and this can be done in a very weak logical system such as Primitive Recursive Arithmetic (PRA). On the other side, in the absolute case, consistency of a formal system is reduced to the consistency of the language of discourse, which, by G\"odel incompleteness theorem, if actually consistent, necessarily has to be as strong as the modelled formal system.}, that is as an interpretation in the language of discourse, which we could call in term the ``metalanguage'' (i.e. interpreting $A$ as a set $\textbf{A}$ or a relation $=_A$ of the language of discourse and interpreting a derivation $t$ of $\vdash A$ as an expression $|t|$ such that, respectively, $|t| \in \textbf{A}$ or $|t| =_A |t|$). Thus, the comparison between parametricity and realisability includes the case of \emph{reducibility}~\cite{Tait}, which is a semantic form of realisability, also known as \emph{glueing} in the context of category theory. In the same vein, binary parametricity over a language can be seen as defining \emph{logical relations}~\cite{Plotkin} over itself, since logical relations are a binary form of reducibility.

@@

Parametricity, like realisability, reducibility, and logical relations, was most commonly handled in an \emph{indexed} form, that is, using a type-theoretic language, considering predicates of type $A \rightarrow \Prop$, or families of type $A \rightarrow \MySet$, or relations of type $A \times A \rightarrow \Prop$, or relevant relations -- sometimes called \emph{correspondences} -- of type $A \times A \rightarrow \MySet$, where $\Prop$ and $\MySet$ are types representing, that is classifying -- in category theory terminology --, propositions or sets, respectively. Along the indexed-fibered correspondence, that is along the equivalence between the types $A \rightarrow \Prop$ and $\Sigma B:\MySet. (B \stackrel{inj}{\rightarrow} A)$ (proof-irrelevant case) and $A \rightarrow \MySet$ and $\Sigma B:\MySet. (B \rightarrow A)$ (proof-relevant case), for $A:\MySet$, everything told in the indexed form can morally also be told in \emph{fibered} form. Such fibered approach to parametricity was for instance considered in Ghani \cite{Ghani...}, who also suggested to iterate parametricity.

@@

The internalisation of parametricity in unary indexed form was studied by \cite{Moulin,Bernardy} directly leading to a type theory of iterated parametricity. Such type theory could be interpreted in a ``unary'' form of cubical set~\cite{BernardyCoquandMoulin}, thus implicitly in fibered form since, in a binary cubical set (resp. in a unary cubical set and simplicial set as well), a set of segments over two points, that is a relevant relation, is given by first and second projection rather than as a function to $\MySet$ (resp. a family is given as a \emph{total space} projected over a set of points $A:\MySet$ rather than a family of type $A \rightarrow \MySet$. This connection between iterated binary parametricity in fibered form and cubical sets includes an identification of degeneracies and reflexivity proofs, which can also be found in Johann and Sojakova~\cite{JohannSojakova}. In particular, it became clear at this point that iterated binary parametricity (without reflexivity) produces semi-cubical sets while it was also observed\footnote{The authors learned it from Hugo Moeneclaye and Thorsten Altenkirch.} that iterated unary parametricity produces augmented semi-simplicial sets.

$$
\begin{tabular}{lllllll}
                           &                            &                                 &                   & structure & kind of semantics \\
 \hline
 \multirow{6}{1cm}{unary}  & \multirow{3}{1.5cm}{indexed} & \multirow{2}{2cm}{non-iterated} & proof-irrelevant  & predicate &        reducibility \\
                           &                            &                                 & relevant          & family & \\
                           &                            & \multicolumn{2}{l}{iterated}                        & \emph{this work}       & \\
                           & \multirow{3}{1.5cm}{fibered} & \multirow{2}{2cm}{non-iterated} & proof-irrelevant  & subobject & \\
                           &                            &                                 & relevant          & fibration & \\
                           &                            & \multicolumn{2}{l}{iterated}                        & augmented semi-simplicial set & \\
 \hline
 \multirow{6}{1cm}{binary} & \multirow{3}{1.5cm}{indexed} & \multirow{2}{2cm}{non-iterated} & proof-irrelevant  & relation & logical relation \\
                           &                            &                                 & relevant          & correspondence (relevant relation) & \\
                           &                            & \multicolumn{2}{l}{iterated}                        & \emph{this work}       & \\
                           & \multirow{3}{1.5cm}{fibered} & \multirow{2}{2cm}{non-iterated} & proof-irrelevant  & fibered relation ? & \\
                           &                            &                                 & relevant          & span & & \\
                           &                            & \multicolumn{2}{l}{iterated}                        & semi-cubical set & \\
 \hline
\end{tabular}
$$

In a previous paper~\citep{hr25}, we exploited this correspondence to construct at the same time augmented semi-simplicial and semi-cubical sets in \emph{indexed form} as a special case of iterated parametricity. That is, instead of the following illustration of the semi-cubical case (upto faces identities),

\begin{equation*}
  \begin{tikzcd}
    X_0: \U & X_1: \U \arrow[l, "\partial^L" description, shift left=2] \arrow[l, "\partial^R" description, shift right=2] & X_2: \U \arrow[l, "\partial^{L\kstar}" description, shift left=6] \arrow[l, "\partial^{R\kstar}" description, shift left=2] \arrow[l, "\partial^{\kstar L}" description, shift right=2] \arrow[l, "\partial^{\kstar R}" description, shift right=6] & \ldots
  \end{tikzcd}
\end{equation*}
we consider a family of relevant relations as shown below:
\begin{equation*}
  \begin{array}{lll}
    E_0 & :               \U                                                                          \\
    E_1 & :               E_0 \times X_0 \rightarrow  \U                                              \\
    E_2 & : \Pi a b c d.\,  E_1(a,b) \times E_1 (c,d) \times E_1(a,c) \times E_1 (b,d) \rightarrow \U \\
    \ldots
  \end{array}
\end{equation*}
